---
title: "DEG btw desmoid and scar-pathway analysis"
output: html_notebook
---

## Load R packages and dependecies
```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(GSVA)
library(clusterProfiler)
library(org.Hs.eg.db)
```


## Load the data
The original file: result/2024-02/02.DDx-Desmoid_vs_Scar.csv
```{r echo=TRUE}
ds <- read.csv("./Folder_for_GitHub/01.DDx-Desmoid_vs_Scar.csv")
```


## Preparing the data for GSEA
```{r echo=TRUE}
ds %>% filter(log2FoldChange >0.2 & padj <0.05 & !is.na(entrez) & baseMean >500) %>% 
  pull(entrez) %>% as.character() -> DESMOID.high

ds %>% filter(log2FoldChange < -0.2 & padj <0.05 & !is.na(entrez)& baseMean >500) %>% 
  pull(entrez) %>% as.character() -> SCAR.high

ds %>% filter(baseMean >500) %>% arrange(desc(log2FoldChange)) %>%
  filter(!is.na(entrez)) %>% dplyr::select(entrez, log2FoldChange) -> geneList.temp
geneList.temp %>% pull(log2FoldChange) -> geneList
names(geneList) <- geneList.temp$entrez
geneList.dedup <- geneList[!duplicated(names(geneList))==TRUE]
```


## DESMOID: GRA using GO database : cellular component
```{r}
ego <- enrichGO(gene          = DESMOID.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
# goplot(ego)  
ego %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## DESMOID: GRA using GO database : Biological Process
```{r}
ego <- enrichGO(gene          = DESMOID.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego@result %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## DESMOID: GRA using GO database : Molecular function
```{r}
ego <- enrichGO(gene          = DESMOID.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego@result %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## SCAR: GRA using GO database : cellular function 
```{r}
ego <- enrichGO(gene          = SCAR.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
# goplot(ego)  
ego %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## SCAR: GRA using GO database : Biological Process 
```{r}
ego <- enrichGO(gene          = SCAR.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## SCAR: GRA using GO database : Mological function 
```{r}
ego <- enrichGO(gene          = SCAR.high,
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego %>% as.data.frame() -> result1.1
Fit <- plot(barplot(ego, showCategory = 20))
print(Fit)
```

## DESMOID: GSEA using GO database : cellular component 
```{r}
# https://support.bioconductor.org/p/99810/
gsea.cc <- gseGO(geneList     = geneList.dedup,
              OrgDb        = org.Hs.eg.db,
              ont          = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              nPermSimple  = 1000000,
              verbose      = FALSE)
# goplot(gsea.cc)   
```


```{r}
gsea.cc@result %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
Idx <- which(gsea.cc$Description == "contractile fiber")
gseaplot(gsea.cc, by = "all", title = gsea.cc$Description[Idx], geneSetID = Idx, color.line="black")
```
GO:0043292 contractile fiber: NES:1.805612 p_val=1.000000e-10 p.adjust=3.450000e-09 qval=2.342105e-09



## DESMOID: GSEA using GO database : biological process 
```{r}
gsea.bp <- gseGO(geneList     = geneList.dedup,
                 OrgDb        = org.Hs.eg.db,
                 ont          = "BP",
                 minGSSize    = 100,
                 maxGSSize    = 500,
                 pvalueCutoff = 0.05,
                 nPermSimple  = 1000000,
                 verbose      = FALSE)
```

```{r}
gsea.bp@result %>% as.data.frame() %>% arrange(desc(NES)) -> result1.1
pdf("./GSEA/gsea.bp_striated_muscle_contraction.pdf")
Idx <- which(gsea.bp$Description == "striated muscle contraction")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0006941, striated muscle contraction: NES:1.799614,  p_val=5.972635e-09, p.adjust=2.945452e-07, qval=2.011835e-07

```{r}
pdf("./GSEA/gsea.bp_leukocyte mediated immunity.pdf")
Idx <- which(gsea.bp$Description == "leukocyte mediated immunity")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0002443, leukocyte mediated immunity, NES:-2.276160, p.val=1.000000e-10, p.adjust=1.041111e-08, qval=7.111111e-09



```{r}
pdf("./GSEA/gsea.bp_lymphocyte mediated immunity.pdf")
Idx <- which(gsea.bp$Description == "lymphocyte mediated immunity")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0002449, lymphocyte mediated immunity,NES:-2.144630, p.val=4.900061e-10, p.adjust=4.012698e-08, qval=2.740797e-08


```{r}
pdf("./GSEA/gsea.bp_tumor necrosis factor production.pdf")
Idx <- which(gsea.bp$Description == "tumor necrosis factor production")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0032640, tumor necrosis factor production, NES:-2.125325, p.val=5.030917e-08, p.adjust=1.745915e-06, q.val=1.192514e-06

```{r}
pdf("./GSEA/gsea.bp_activation of immune response.pdf")
Idx <- which(gsea.bp$Description == "activation of immune response")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0002253 activation of immune response, NES:-2.048426, pval=1.000000e-10, p.adjust=1.041111e-08, qval=7.111111e-09

```{r}
pdf("./GSEA/gsea.bp_adaptive immune response.pdf")
Idx <- which(gsea.bp$Description == "adaptive immune response")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0002250 adaptive immune response NES:-1.991511, pval=0.0000000001, p.adjust=1.041111e-08, qval=7.111111e-09


```{r}
pdf("./GSEA/gsea.bp_positive regulation of T cell activation.pdf")
Idx <- which(gsea.bp$Description == "positive regulation of T cell activation")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0050870 positive regulation of T cell activation NES:-1.821711 pval=1.790241e-06 p.adust=2.713274e-05 qval=1.853250e-05


```{r}
pdf("./GSEA/gsea.bp_T cell proliferation.pdf")
Idx <- which(gsea.bp$Description == "T cell proliferation")
gseaplot(gsea.bp, by = "all", title = gsea.bp$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0042098 T cell proliferation NES:-1.680865 pval=1.798944e-04 p.adjust=0.0013272521 qval=9.065543e-04


## DESMOID: GSEA using GO database : Molecular Function 
```{r}
gsea.mf <- gseGO(geneList     = geneList.dedup,
                 OrgDb        = org.Hs.eg.db,
                 ont          = "MF",
                 minGSSize    = 100,
                 maxGSSize    = 500,
                 pvalueCutoff = 0.05,
                 nPermSimple  = 1000000,
                 verbose      = FALSE)
```

```{r}
gsea.mf@result %>% as.data.frame() %>% arrange(desc(NES)) -> result1.1
pdf("./GSEA/gsea.bp_gated channel activity.pdf")
Idx <- which(gsea.mf$Description == "gated channel activity")
gseaplot(gsea.mf, by = "all", title = gsea.mf$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GO:0022836 gated channel activity NES:1.797640 pval=4.549420e-09 p.adjust=2.957123e-07 qval=2.466264e-07


## GSEA analysis using msigdb
##### https://cran.r-project.org/web/packages/msigdbr/vignettes/msigdbr-intro.html
```{r}
library(msigdbr)
msigdbr_show_species()
m_df <- msigdbr(species = "Homo sapiens")
head(m_df, 2) %>% as.data.frame
unique(m_df$gs_subcat)
```

### ALL msigdb gene sets 
```{r}
m_t2g <- m_df %>% dplyr::select(gs_name, entrez_gene)
em <- enricher(DESMOID.high, TERM2GENE=m_t2g)
em %>% as.data.frame() -> result1.1
```


```{r}
gsea.all <- GSEA(geneList.dedup, TERM2GENE = m_t2g, nPermSimple = 1000000)
```

```{r}
gsea.all %>% as.data.frame() -> result1.1
pdf("./GSEA/gsea.bp_GSE15659_RESTING_TREG_VS_NONSUPPRESSIVE_TCELL_UP.pdf")
Idx <- which(gsea.all$Description == "GSE15659_RESTING_TREG_VS_NONSUPPRESSIVE_TCELL_UP")
gseaplot(gsea.all, by = "all", title = gsea.all$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GSE15659_RESTING_TREG_VS_NONSUPPRESSIVE_TCELL_UP NES:1.609330, pval=0.0001621542, p.adjust=0.004018443 qval=0.003375783

### Hall mark gene sets
```{r}
H_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)
gsea.h <- GSEA(geneList.dedup, TERM2GENE = H_t2g)
gsea.all %>% as.data.frame() -> result1.1
```


### C2 Curated gene sets  CGP chemical and genetic perturbations
```{r}
C2CGP_t2g <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CGP") %>%
  dplyr::select(gs_name, entrez_gene)
gsea.c2cgp <- GSEA(geneList.dedup, TERM2GENE = C2CGP_t2g, nPermSimple = 1000000)
gsea.c2cgp %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```


### C5 ontology gene sets  HPO: human phenotype ontology
```{r}
C5HPO_t2g <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "HPO") %>%
  dplyr::select(gs_name, entrez_gene)
gsea.c5hpo <- GSEA(geneList.dedup, TERM2GENE = C5HPO_t2g, nPermSimple = 1000000)
gsea.c5hpo %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```


### C3 regultatory target gene sets TFT: transcription factor targets 
TFT_LEGACY subset of TFT (browse 610 gene sets)
```{r}
C3TF_legacy_t2g <- msigdbr(species = "Homo sapiens", category = "C3", subcategory = "TFT:TFT_Legacy") %>% 
  dplyr::select(gs_name, entrez_gene)
gsea.c3tft_leg <- GSEA(geneList.dedup, TERM2GENE = C3TF_legacy_t2g, nPermSimple = 1000000)
gsea.c3tft_leg %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```

### C3 regultatory target gene sets TFT: transcription factor targets 
GTRD subset of TFT (browse 505 gene sets)
```{r}
C3TF_GTRD_t2g <- msigdbr(species = "Homo sapiens", category = "C3", subcategory = "TFT:GTRD") %>% 
  dplyr::select(gs_name, entrez_gene)
gsea.c3tft_gtrd <- GSEA(geneList.dedup, TERM2GENE = C3TF_GTRD_t2g, nPermSimple = 1000000)
gsea.c3tft_gtrd %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```

### C6: oncogenic signature gene sets (browse 189 gene sets)
```{r}
C6_t2g <- msigdbr(species = "Homo sapiens", category = "C6") %>% 
  dplyr::select(gs_name, entrez_gene)
gsea.c6 <- GSEA(geneList.dedup, TERM2GENE = C6_t2g, nPermSimple = 1000000)
gsea.c6 %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```
```{r}
Idx <- which(gsea.c6$Description == "KRAS.KIDNEY_UP.V1_UP")
gseaplot(gsea.c6, by = "all", title = gsea.c6$Description[Idx], geneSetID = Idx, color.line="black")
```



### C7: immunologic signature gene sets ImmuneSigDB subset of C7 (browse 4872 gene sets)
```{r}
C7_t2g<- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)
gsea.c7 <- GSEA(geneList.dedup, TERM2GENE = C7_t2g, nPermSimple = 1000000)
```

```{r}
gsea.c7 %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
pdf("./GSEA/gsea.bp_GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_IRES_GFP_UP.pdf")
Idx <- which(gsea.c7$Description == "GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_IRES_GFP_UP")
gseaplot(gsea.c7, by = "all", title = gsea.c7$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_IRES_GFP_UP NES:1.807670 pval=4.308232e-06 p.adjust=0.000255972 qval=0.0002302885

```{r}
pdf("./GSEA/gsea.bp_GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_FUSION_GFP_UP.pdf")
Idx <- which(gsea.c7$Description == "GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_FUSION_GFP_UP")
gseaplot(gsea.c7, by = "all", title = gsea.c7$Description[Idx], geneSetID = Idx, color.line="black")
dev.off()
```
GSE37605_TREG_VS_TCONV_C57BL6_FOXP3_FUSION_GFP_UP NES:1.650602 pval=2.527618e-04 p.adjust=0.006282937 qval=0.0056525255

### C8: cell type signature gene sets (browse 830 gene sets)
```{r}
C8_t2g<- msigdbr(species = "Homo sapiens", category = "C8") %>% 
  dplyr::select(gs_name, entrez_gene)
gsea.c8 <- GSEA(geneList.dedup, TERM2GENE = C8_t2g, nPermSimple = 1000000)
gsea.c8 %>% as.data.frame() %>% arrange(desc(NES))-> result1.1
```


```{r}
save(gsea.all, gsea.bp, gsea.c2cgp, gsea.c3tft, gsea.c3tft_gtrd, gsea.c3tft_leg,
     gsea.c5hpo, gsea.c6, gsea.c7, gsea.c8, gsea.cc, gsea.h, gsea.mf, 
     file="./Desmoid_scar_GSEA.Rda")
```



